cardinalR: Generating interesting high-dimensional data structures

### 2025-05-04 paper-cardinalR.Rmd
Added grided sphere description

@@

>>>>

The `gen_unifsphere(n, p, r)` function generates a $p$-dimensional dataset of $n$ observations distributed approximately uniformly on the surface of a $3\text{-}D$ sphere of radius $r$, with additional Gaussian noise dimensions added when $p > 3$.

Each observation lies on the surface of a sphere in $3\text{-}D$, constructed by generating $u \sim U(-1, 1)$ which represents the cosine of the polar angle $\phi$ and $\theta \sim U(0, 2\pi)$ which represents the azimuthal angle.

The corresponding Cartesian coordinates are calculated as;

$$
X_1 = r \cdot \sqrt{1 - u^2} \cdot \cos(\theta),
$$
$$
X_2 = r \cdot \sqrt{1 - u^2} \cdot \sin(\theta),
$$
$$
X_3 = r \cdot u,
$$

which gives points uniformly distributed on the surface of a $3\text{-}D$ sphere (not within).

For $p > 3$, additional dimensions $X_4$ to $X_p$ are generated as low-variance Gaussian noise:

$$
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 4, \dots, p.
$$

This extends the $3\text{-}D$ spherical manifold into $p$-dimensional space with orthogonal Gaussian noise, preserving the spherical structure while allowing for testing in higher-dimensional settings.

@@

The `gen_gridedsphere(n, p)` function generates a $p$-dimensional dataset of approximately $n$ observations evenly distributed on the surface of a $3\text{-}D$ unit sphere, with optional Gaussian noise dimensions when $p > 3$.

The base structure consists of a $3\text{-}D$ spherical surface created using a regular grid in spherical coordinates, where $\theta \in [0, 2\pi]$ represents the azimuthal angle (longitude) and $\phi \in [0, \pi]$ denotes the polar angle (co-latitude).

The number of grid steps along each dimension is determined by factoring $n$ into two approximately equal integers via `gen_nproduct(n, p = 2)`.

Each point on the sphere is computed using the spherical-to-Cartesian transformation:

$$
X_1 = \sin(\phi) \cdot \cos(\theta),
$$
$$
X_2 = \sin(\phi) \cdot \sin(\theta),
$$
$$
X_3 = \cos(\phi).
$$

This forms a structured grid of points on the unit sphere, ideal for studying how regular geometric manifolds are transformed under nonlinear dimensionality reduction.

For $p > 3$, the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions:

$$
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 4, \dots, p.
$$

These dimensions represent low-variance orthogonal noise, preserving the spherical shape while testing robustness in high-dimensional settings.
